---
title: "Analysis from 2020-2024"
author: "Anthony Fernandez"
date: "2025-06-05"
output: html_document
---

```{r message=FALSE, warning=FALSE}
library(readr)
library(dplyr)

# Read in each year's data and add a Year column
tc_2020 <- read_csv("team_comparison_20.csv") %>% mutate(Year = 2020)
tc_2021 <- read_csv("team_comparison_21.csv") %>% mutate(Year = 2021)
tc_2022 <- read_csv("team_comparison_22.csv") %>% mutate(Year = 2022)
tc_2023 <- read_csv("team_comparison_23.csv") %>% mutate(Year = 2023)
tc_2024 <- read_csv("team_comparison_24.csv") %>% mutate(Year = 2024)

# Combine into one data frame
team_comparison_all <- bind_rows(tc_2020, tc_2021, tc_2022, tc_2023, tc_2024)

```

```{r}
library(ggplot2)
library(dplyr)

ggplot(team_comparison_all, aes(x = Year, y = BA_Difference, group = batting_team, color = batting_team)) +
  geom_line(alpha = 0.6) +
  geom_point() +
  labs(
    title = "Momentum Effect by Team Over Time (2020–2024)",
    x = "Season",
    y = "Difference in Batting Average (Momentum - No Momentum)",
    color = "Team"
  ) +
  theme_minimal(base_size = 13)

```

```{r}
team_comparison_all %>%
  group_by(Year) %>%
  summarise(Avg_BA_Diff = mean(BA_Difference, na.rm = TRUE)) %>%
  ggplot(aes(x = Year, y = Avg_BA_Diff)) +
  geom_line(color = "#2c3e50", size = 1.2) +
  geom_point(size = 3, color = "#e67e22") +
  labs(
    title = "League-Wide Average Momentum Batting Average Difference",
    y = "Avg. BA Difference",
    x = "Year"
  ) +
  theme_minimal(base_size = 13)

```

```{r}
library(tidyr)

trend_summary <- team_comparison_all %>%
  group_by(batting_team) %>%
  summarise(trend = lm(BA_Difference ~ Year)$coefficients[2]) %>%
  arrange(desc(trend)) %>%
  slice(1:5)

top_teams <- trend_summary$batting_team

team_comparison_all %>%
  filter(batting_team %in% top_teams) %>%
  ggplot(aes(x = Year, y = BA_Difference, color = batting_team)) +
  geom_line(size = 1.1) +
  geom_point(size = 2) +
  labs(
    title = "Top 5 Teams Increasing in Momentum Effect",
    y = "BA Difference",
    x = "Year"
  ) +
  theme_minimal(base_size = 13)

```


```{r}
library(cluster)
library(factoextra)
library(tibble)


pivoted <- team_comparison_all %>%
  select(batting_team, Year, BA_Difference) %>%
  tidyr::pivot_wider(names_from = Year, values_from = BA_Difference)

# Replace NAs with column means for clustering
pivoted_clean <- pivoted %>%
  column_to_rownames("batting_team") %>%
  mutate(across(everything(), ~ifelse(is.na(.), mean(., na.rm = TRUE), .)))

# Run K-means
set.seed(42)
km <- kmeans(pivoted_clean, centers = 5, nstart = 25)

fviz_cluster(km, data = pivoted_clean) +
  labs(title = "Clustering Teams by Momentum Patterns (2020–2024)")

```

### Cluster Narratives

- **Cluster 1 – Upward Momentum Growers**: Slightly negative in 2020, improving every year through 2024.
- **Cluster 2 – Steady, Modest Momentum Teams**: Mildly positive and consistent.
- **Cluster 3 – Inconsistent or Rebounding Momentum Teams**: Big dip in 2020, mixed recovery.
- **Cluster 4 – Elite Momentum Maximizers**: Strongest and steadily rising BA differences.
- **Cluster 5 – Front-Loaded Momentum Teams**: Huge 2020 momentum boost, flatter since.