---
title: "Momentum-in-the-MLB-2023"
author: "Anthony Fernandez"
date: "2025-06-05"
output: html_document
---


```{r message=FALSE, warning=FALSE}
devtools::install_github("billpetti/baseballr")

library(baseballr)
library(dplyr)
library(purrr)
library(lubridate)
library(tidyverse)


# Load required packages
#install.packages("data.table")  # Only run if needed
library(data.table)
library(dplyr)
library(purrr)

```

```{r message=FALSE, warning=FALSE}


# Safe downloader
get_statcast_data_weekly <- function(start_date, end_date) {
  url <- paste0(
    "https://baseballsavant.mlb.com/statcast_search/csv?all=true",
    "&hfPT=&hfAB=&hfBBT=&hfPR=&hfZ=&stadium=&hfBBL=&hfNewZones=&hfGT=R%7CPO%7CS%7C",
    "&hfC&hfSit=&hfOuts=&opponent=&pitcher_throws=&batter_stands=",
    "&hfSA=&player_type=batter&hfInfield=&team=&position=&hfOutfield=",
    "&hfRO=&home_road=&game_date_gt=", start_date,
    "&game_date_lt=", end_date,
    "&group_by=name&sort_col=pitches&player_event_sort=h_launch_speed",
    "&sort_order=desc&type=details"
  )
  
  message(paste("Downloading:", start_date, "to", end_date))
  tryCatch({
    fread(url, showProgress = FALSE, fill = TRUE)
  }, error = function(e) {
    message("Failed:", e$message)
    NULL
  })
}

```

```{r message=FALSE}
# Define weekly ranges for 2023 regular season
dates_2023 <- data.frame(
  start = c("2023-03-30", "2023-04-08", "2023-04-15", "2023-04-22", "2023-04-29",
            "2023-05-06", "2023-05-13", "2023-05-20", "2023-05-27",
            "2023-06-03", "2023-06-10", "2023-06-17", "2023-06-24",
            "2023-07-01", "2023-07-08", "2023-07-15", "2023-07-22", "2023-07-29",
            "2023-08-05", "2023-08-12", "2023-08-19", "2023-08-26",
            "2023-09-02", "2023-09-09", "2023-09-16", "2023-09-23", "2023-09-30"),
  end = c("2023-04-07", "2023-04-14", "2023-04-21", "2023-04-28", "2023-05-05",
          "2023-05-12", "2023-05-19", "2023-05-26", "2023-06-02",
          "2023-06-09", "2023-06-16", "2023-06-23", "2023-06-30",
          "2023-07-07", "2023-07-14", "2023-07-21", "2023-07-28", "2023-08-04",
          "2023-08-11", "2023-08-18", "2023-08-25", "2023-09-01",
          "2023-09-08", "2023-09-15", "2023-09-22", "2023-09-29", "2023-10-01")
)

# Updated function using the exported method
get_statcast_data_weekly <- function(start_date, end_date) {
  message("Downloading: ", start_date, " to ", end_date)
  baseballr::scrape_statcast_savant_batter(start_date, end_date)
}

# Run the download loop
#statcast_2023 <- purrr::map2_dfr(dates_2023$start, dates_2023$end, get_statcast_data_weekly)

# Optional: Save result
#saveRDS(statcast_2023, "statcast_2023_full.rds")

statcast_2023 <- readRDS("statcast_2023_full.rds")

```




```{r}
#summary(statcast_2023)

cols_to_keep <- c(
  "game_pk",             # Game ID
  "game_date",           # Date of game
  "at_bat_number",       # Sequential number of PA in game
  "inning",              # Inning
  "inning_topbot",       # Top or Bottom of inning
  "batter",              # Batter ID
  "player_name",         # Batter name
  "events",              # Event result (e.g. 'single', 'groundout')
  "description",         # Useful for identifying completed PAs
  "home_team",           # Home team
  "away_team",           # Away team
  "bat_score",           # Score at the time
  "post_bat_score",      # Score after PA
  "home_score", "away_score", # Full context
  "pitch_number"         # Needed to make sure we're only using the final pitch of the PA
)

statcast_2023_clean <- statcast_2023 %>%
  select(all_of(cols_to_keep)) %>%
  filter(!is.na(events))  # Remove non-PA rows (e.g. pickoffs, warmups)

# Check result
glimpse(statcast_2023_clean)

unique(statcast_2023$events)


```

```{r}
# Keep only the final pitch of each at-bat (where event is not NA and last pitch_number per at_bat)
library(dplyr)

statcast_pa_2023 <- statcast_2023_clean %>%
  filter(!is.na(events), events != "") %>%
  group_by(game_pk, at_bat_number) %>%
  filter(pitch_number == max(pitch_number)) %>%
  ungroup()

```

```{r}
# Sort by game, inning, top/bot, and at-bat
statcast_pa_2023 <- statcast_pa_2023 %>%
  arrange(game_pk, inning, inning_topbot, at_bat_number) %>%
  group_by(game_pk, inning, inning_topbot) %>%
  mutate(
    prev_event = lag(events),
    prev_batter = lag(batter),
    momentum = ifelse(prev_event %in% c("single", "double", "triple", "home_run", "sac_fly"), 1, 0)
  ) %>%
  ungroup()

```

```{r}
# Calculate batting average with vs. without momentum
momentum_ba <- statcast_pa_2023 %>%
  filter(events %in% c(
    "single", "double", "triple", "home_run",
    "field_out", "strikeout", "grounded_into_double_play",
    "double_play", "strikeout_double_play",
    "fielders_choice", "force_out"
  )) %>%
  mutate(hit = ifelse(events %in% c("single", "double", "triple", "home_run"), 1, 0)) %>%
  group_by(momentum) %>%
  summarise(
    PA = n(),
    Hits = sum(hit),
    Batting_Average = mean(hit)
  )

print(momentum_ba)


```





############################################
########Break Down by Team##################
############################################

```{r}
statcast_pa_2023 <- statcast_pa_2023 %>%
  mutate(batting_team = ifelse(inning_topbot == "Top", away_team, home_team))

statcast_pa_2023 <- statcast_pa_2023 %>%
  mutate(hit = ifelse(events %in% c("single", "double", "triple", "home_run"), 1, 0))

```

```{r}
library(dplyr)

team_momentum_ba <- statcast_pa_2023 %>%
 filter(events %in% c(
    "single", "double", "triple", "home_run",
    "field_out", "strikeout", "grounded_into_double_play",
    "double_play", "strikeout_double_play",
    "fielders_choice", "force_out"
  )) %>%
  group_by(batting_team, momentum) %>%
  summarise(
    PA = n(),
    Hits = sum(hit),
    Batting_Average = mean(hit),
    .groups = "drop"
  )

```

```{r}
library(tidyr)

team_comparison_23 <- team_momentum_ba %>%
  pivot_wider(
    names_from = momentum,
    values_from = c(PA, Hits, Batting_Average),
    names_prefix = "momentum_"
  ) %>%
  mutate(
    BA_Difference = Batting_Average_momentum_1 - Batting_Average_momentum_0
  ) %>%
  arrange(desc(BA_Difference))

# View top teams by momentum boost
print(team_comparison_23)

#write.csv(team_comparison_23, "team_comparison_23.csv", row.names = FALSE)

```

```{r}
# Total hits in cleaned dataset
sum(statcast_pa_2023$hit, na.rm = TRUE)

# Total hits from your momentum breakdown
sum(team_comparison_23$Hits_momentum_0, na.rm = TRUE) + 
sum(team_comparison_23$Hits_momentum_1, na.rm = TRUE)

```


The reason the total hits from the momentum breakdown (Hits_momentum_0 + Hits_momentum_1) don't match official MLB hit totals is because my analysis only includes plate appearances with specific qualifying events (like hits, outs, and double plays), and excludes others like walks, hit-by-pitches, and sac bunts. Additionally, momentum can only be calculated when there's a valid previous batter, so the first batter of each inning or game may be excluded. However, since my total hits from the filtered dataset match the combined momentum hits exactly (36,855), my momentum analysis is internally consistent and correctly reflects the subset of data I chose to study.


```{r}
# Filter valid at-bats and create the hit column
valid_pas <- statcast_pa_2023 %>%
  filter(events %in% c(
    "single", "double", "triple", "home_run",
    "field_out", "strikeout", "grounded_into_double_play",
    "double_play", "strikeout_double_play",
    "fielders_choice", "force_out"
  )) %>%
  mutate(hit = ifelse(events %in% c("single", "double", "triple", "home_run"), 1, 0))

# Run the t-test
t_test_result <- t.test(hit ~ momentum, data = valid_pas)

print(t_test_result)

```


```{r}
library(ggplot2)

ggplot(team_comparison_23 %>% filter(!is.na(batting_team)), 
       aes(x = reorder(batting_team, BA_Difference), y = BA_Difference, fill = BA_Difference > 0)) +
  geom_col(width = 0.7, color = "white", show.legend = FALSE) +
  scale_fill_manual(values = c("FALSE" = "#e74c3c", "TRUE" = "#2ecc71")) +
  geom_text(aes(label = sprintf("%+d pts", round(BA_Difference * 1000))), 
            hjust = ifelse(team_comparison_23$BA_Difference > 0, -0.15, 1.1), 
            size = 3.5, color = "black") +
  coord_flip() +
  labs(
    title = "Which Teams Ride the Momentum?",
    subtitle = "Difference in Batting Average When Previous Batter Got a Hit or Sac Fly (2023)",
    x = "Team",
    y = "Momentum Batting Avg. Difference (in points)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "gray30", size = 12),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(face = "bold")
  )

```